% experiments/exp2_chain_recipe/abstract_v3.tex
%
% Replacement abstract for draft.tex incorporating the v14/v15
% findings of section7_v14_findings.tex and
% section8_recipe_revised.tex, plus the callback-aware standalone
% eval methodology of section_eval_callback.tex.
%
% This file is meant to be \input'd in place of the original
% \begin{abstract}...\end{abstract} block of draft.tex. It does not
% modify draft.tex.

\begin{abstract}
We study the training recipe needed to make a parity-preserving
recurrent memory --- the \emph{Memory Residuals} primitive ---
useful for long-horizon multi-session dialogue recall, across three
iterations of empirical revision. The first iteration gives a
four-piece recipe (contextual extract source, mixed conversational
corpus, negative-chain contrastive supervision, soft router init)
and a sharp negative result on the strict $\pm 32$ parity init
(\S\ref{sec:negative_result}), which we show is operationally
untrainable under any contrastive ramp because it places the
routing softmax beyond bf16-representable gradient. A second
iteration adds a supervised writer warmup; a v14 ablation on a
controlled synthetic callback corpus (D4v2) retires it. On D4v2 the
warmup imprints a ``compress everything'' policy on the writer
that, post-anneal, locks the router into a recent-bias fixed point
with $\amem^{\text{mean}}{=}\tau{=}0.05$ exactly and \emph{negative}
evidence-on-callback lift. The revised \emph{three}-piece recipe
(\S\ref{sec:revised_recipe}) is input-RMSNorm on the extractor,
\emph{no} writer warmup, and alpha-floor $+$ multi-negative
InfoNCE; on a frozen Qwen3-0.6B backbone it is the first recipe in
the paper that produces sustained positive callback evidence-lift.
We further document a methodological finding (\S\ref{sec:eval_cb}):
standard chain-eval averaging dilutes the callback-positioned
signal across roughly four sessions of filler-and-evidence, driving
its signal-to-noise below one. A \emph{callback-aware} standalone
eval --- restricted to the callback session and to the annotated
chain-callback-position tokens within it, with the $\dnm$-floor
measured against an evidence-redacted $\Mc$ --- recovers a
$+1.4$~nat memory-on-callback effect on the v14k step-$3000$
checkpoint
($\dnm^{\text{cb}}{=}{+1.44}$,
 $\dsh^{\text{cb}}{=}{+0.054}$,
 $\text{evidence\_lift}{=}{+0.071}$)
that the standard full-window eval averages to noise
($\dnm{=}{-0.085}$ on the same checkpoint). We close with an honest
negative on joint training: with the backbone unfrozen at $0.6$B,
the bare backbone learns the D4v2 template directly and the memory
signal disappears regardless of recipe. The $1.7$B headline numbers
under the revised recipe (Table~\ref{tab:revised_headline}) are
pending the v15e/v15f cells currently in flight.
\end{abstract}
