% experiments/exp2_long_horizon_recipe/draft.tex
%
% A training recipe for long-horizon dialogue recall on Memory Residuals.
%
% Companion to experiments/exp1_drop_in_primitive/manuscript.tex
% (drop-in primitive paper). After the first review pass that
% paper was deemed too narrow to stand alone, so the architectural
% primitive is folded into this manuscript as the "Background /
% Architectural primitive" section -- the rest of the paper is the
% training-recipe contribution.
%
\documentclass{article}

\usepackage[preprint]{neurips_2024}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{hyperref}
\usepackage{url}
\usepackage{booktabs}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{microtype}
\usepackage{xcolor}
\usepackage{graphicx}
\usepackage{multirow}
\usepackage{makecell}
\usepackage{algorithm}
\usepackage{algpseudocode}

\newcommand{\todo}[1]{\textcolor{red}{\textbf{[TODO: #1]}}}
\newcommand{\mhyph}{\text{--}}
\newcommand{\Mc}{M_{\text{c}}}
\newcommand{\Min}{M_{\text{in}}}
\newcommand{\Mj}{M_{\text{judge}}}
\newcommand{\Mnew}{M_{\text{new}}}
\newcommand{\dnm}{\Delta_{\text{nm-m}}}
\newcommand{\dsh}{\Delta_{\text{sh-m}}}
\newcommand{\dor}{\Delta_{\text{or-m}}}
\newcommand{\CEmem}{\mathrm{CE}_{\text{mem}}}
\newcommand{\CEnomem}{\mathrm{CE}_{\text{nomem}}}
\newcommand{\CEsh}{\mathrm{CE}_{\text{shuffle}}}
\newcommand{\CEor}{\mathrm{CE}_{\text{oracle}}}
\newcommand{\amem}{\alpha_{\text{mem}}}

\title{A training recipe for long-horizon dialogue recall on\\
       Memory Residuals: contextual extraction, mixed corpus,\\
       contrastive supervision, and the cost of saturating router init}

\author{%
  Anonymous Authors\\
  \texttt{anonymous@example.com}%
}

\begin{document}
\maketitle

\begin{abstract}
We study the training recipe needed to make a parity-preserving
recurrent memory module --- the \emph{Memory Residuals} primitive ---
useful for long-horizon multi-session dialogue recall. The primitive
admits a bit-exact initialisation that produces zero perturbation of
the bare backbone's logits at step~$0$; it also, under na\"ive
next-token training on book corpora, is provably sample-efficient
on history-specificity diagnostics ($\dsh$) at matched compute
relative to a ReZero scalar gate. \emph{However}, that same recipe
fails to transfer to dialogue: standalone evaluation on LoCoMo and
the held-out routing-mass / counterfactual probes both show that the
depth-router never opens a non-trivial channel to memory, with
$\amem \approx 5\!\times\!10^{-4}$ even after $4400$ steps of
optimisation. We trace the failure to four interacting causes ---
embedding-only extract source, book-skewed training mix,
no signal beyond plain NLL, and an init that places the routing
softmax beyond bf16-representable gradient --- and propose a
four-piece recipe to fix each. We then report on the headline
$2\!\times\!2$ ablation (contextual extract $\times$ mixed corpus)
trained under that recipe, and on a key negative result: the strict
parity-preserving init ($\pm 32$) produces a literal numerical-zero
gradient through the router softmax, so the contrastive ramp cannot
escape it.

A second iteration of the recipe, informed by three diagnostic
signatures surfaced during the first-iteration runs, adds three
further training-time pieces on top of the architectural writer
refinement introduced in the position paper~\cite{liu2025memres}:
a supervised writer warmup that decouples the two-stage QKV
subsystem's optimisation from the joint LM objective, a
load-balance auxiliary loss on the router's per-sublayer memory
mass $\amem$ to prevent MoE-style expert collapse of the memory
source under joint training, and a multi-negative InfoNCE
contrastive term that gives the writer a direct discriminative
target beyond the margin-style negative-chain loss of Piece~3.
We establish that each is necessary on dialogue corpora at
$0.6$B scale, and that the combined seven-piece recipe is what
finally yields a router that recruits memory above a target
floor throughout joint training, a writer whose slots remain
orthogonal across the full run, and a judge that makes non-uniform
decisions.
\end{abstract}

\section{Introduction}
\label{sec:intro}

A practical conversational agent should remember a user across
sessions: their name, their preferences, the on-going arc of a
relationship. \emph{Memory Residuals}~\cite{liu2025memres,
duan2025memres} is an architectural primitive aimed at exactly
this regime --- a fixed-size recurrent memory matrix
$\Mc \in \mathbb{R}^{K\times d}$ updated end-to-end through the
language-modelling loss with no retrieval index, no separate memory
controller, and no hand-engineered gating heuristic. The companion
paper~\cite{liu2025drop} establishes that the primitive, on top of a
$0.6$B Qwen3 backbone and trained on book continuations (PG-19 plus
$30$ TV transcripts), is roughly $2\times$ more sample-efficient on
the history-specificity test ($\dsh$, defined below) than its
ReZero-style gate baseline; that the parity-preserving init produces
bit-exact logits to the bare backbone at step $0$; and that on
held-out PG-19 the router opens a non-trivial channel to memory,
with the routing-mass and counterfactual probes both registering
positive but small effects.

The primitive, however, does not by itself transfer to dialogue.
Trained on the book corpus and evaluated standalone on LoCoMo
(a multi-session dialogue benchmark), the same checkpoint shows
$\dsh$ within bootstrap CI of zero. The held-out routing trace shows
$\amem$ at $\sim 5\!\times\!10^{-4}$, only $7.6\times$ off the
analytic init floor of $\exp(-8)/N \approx 6\!\times\!10^{-5}$. The
counterfactual depth-bucketed probe shows held-out NLL effects of
$\Delta(d{=}\mathrm{ALL}) \approx +0.013$~nat that drop into noise
beyond the most-recent prior session.

\paragraph{What this paper contributes.} A training recipe that
transforms the Memory Residuals primitive from \emph{book-only,
margin-effect} to \emph{dialogue-capable, mechanistically active}
on the same backbone and at compute parity. The recipe has four
pieces, each motivated by a specific failure mode of the
primitive's na\"ive training on book data:

\begin{enumerate}\setlength{\itemsep}{0pt}
\item \textbf{Contextual extract source.} Replace the legacy
bag-of-token-embeddings extract input with a hidden state from
mid-stack of the bare backbone, $\texttt{hidden\_<L>}$. The
compressor's distillation target shifts from token identity to
contextualised semantics, which is necessary to bind dialogue
content across sessions whose token surfaces differ.
\item \textbf{Mixed conversational corpus} (PG-19 + TV + MSC) with
explicit per-source weighting. Train on the kind of data we want
to evaluate on; include enough non-dialogue to anchor the LM head.
\item \textbf{Negative-chain contrastive loss}, ramped over the
first $1000$ steps. The contrast forces the matched-chain memory
to beat a distractor chain's memory by a margin on the same
scoring session, giving the router a reason to recruit memory
beyond what plain NLL provides on dialogue.
\item \textbf{Soft, not strict, parity-preserving router init.}
The strict $\pm 32$ bias init is bit-exact at step $0$ but produces
a router softmax gradient on memory that is below bf16
representability. The contrastive loss cannot escape this
saturation. We show this experimentally and recommend the soft
$\pm 4$ init as the pragmatic compromise between init purity and
recipe-trainable optimisation.
\end{enumerate}

The headline experiment is a $2\!\times\!2$ factorial over
$\{\texttt{embed},\,\texttt{hidden\_14}\}\times\{\textsc{PG19+TV},\,
\textsc{PG19+TV+MSC}\}$, with the $\texttt{hidden\_14}\!\times\!\textsc{PG19+TV+MSC}$
cell as the headline. We report $\dnm$, $\dsh$, and held-out routing
trace + counterfactual probes for every cell, with bootstrap CIs.
We also report the negative result on the strict $\pm 32$ init,
which is, in our view, the most actionable finding for follow-on
work because it identifies a previously implicit failure mode of
parity-preserving initialisation in residual-stream routers.

\paragraph{Scope and non-scope.} We report a $0.6$B-class result
on Qwen3. We do not claim $32$k context: the operational scope is
$k$-session windows of $S{=}512$ tokens each, with $k\!\le\!8$
sessions. We hold MSC and LoCoMo as evaluation surfaces only ---
held-out by user mandate from the training corpus mix. We do not
attempt to compete with retrieval-augmented generation on factual
$\mathrm{QA}$ benchmarks; the comparison we run is to a
compute-matched RAG baseline on the dialogue corpora the recipe is
trained for, where the differentiator is abstract callback rather
than verbatim retrieval.

\paragraph{Roadmap.} \S\ref{sec:primitive} folds in the
architectural primitive's definition and its book-only baseline
performance from \cite{liu2025drop}, so this manuscript can stand
on its own. \S\ref{sec:recipe} states the four-piece recipe and
the prior failure mode each piece addresses. \S\ref{sec:setup}
describes the corpus build and training infrastructure.
\S\ref{sec:experiments} reports the $2\!\times\!2$ headline,
RAG-baseline comparisons, callback probe, and routing-trace +
counterfactual diagnostics. \S\ref{sec:negative_result} pins down
the strict-vs-soft init failure mode in detail, including a
bf16-arithmetic argument for why the strict $\pm 32$ init is
operationally untrainable under a contrastive ramp.
\S\ref{sec:second_iteration} presents the diagnostic findings that
the first-iteration recipe, while sufficient to move $\dsh$ off zero
on dialogue, leaves three well-defined training-dynamics failure
signatures standing; and it introduces three additional recipe
pieces (writer warmup, $\amem$ load-balance aux loss, multi-negative
InfoNCE contrastive) that target those signatures individually,
paired with the architectural judge refinement of the position
paper~\cite{liu2025memres}. \S\ref{sec:discussion} discusses
limitations and the interaction of architectural and training-recipe
interventions.

\section{Background: the architectural primitive}
\label{sec:primitive}

This section is a self-contained summary of the primitive defined
and benchmarked in \cite{liu2025drop}. We re-derive only what later
sections depend on; readers interested in the primitive's
sample-efficiency vs.\ scalar-gate baseline should consult that
paper directly.

\paragraph{Memory state.} Each chain (a book, a TV episode arc, or
a multi-session conversation) carries a single fixed-size matrix
$\Mc \in \mathbb{R}^{K\times d}$ with $K{=}128$ slots and $d$ the
backbone hidden size ($1024$ for Qwen3-0.6B). The state is updated
once per session boundary via a two-stage cross-attention:
\begin{align}
\Mnew &= \mathrm{Softmax}\!\left(\tfrac{\Min W_Q^{\text{ext}}\bigl(C_t W_K^{\text{ext}}\bigr)^\top}{\sqrt{d}}\right) C_t W_V^{\text{ext}},
\label{eq:extract}\\
\Mc^{t} &= \mathrm{Softmax}\!\left(\tfrac{\Mj W_Q^{\text{j}}\bigl(P W_K^{\text{j}}\bigr)^\top}{\sqrt{d}}\right) P W_V^{\text{j}},
\quad P = [\Mc^{t-1} \parallel \Mnew] \in \mathbb{R}^{2K\times d},
\label{eq:judge}
\end{align}
with extraction depth $L_E{=}4$ and judge softmax taken across the
$2K$-row dimension. The $2K$-pool softmax is the
\emph{zero-sum forgetting defence}: salient new content displaces
old slots; mundane content lets old content pass through.
$C_t$ is the per-session input to the extractor, and is the
locus of the recipe's Piece~1 below.

\paragraph{Off-sequence read.} At inference into the next session,
every position $x_i$ cross-attends $\Mc$ once to produce a
per-position readout
$m^t = \mathrm{Softmax}(X W_Q^{\text{r}}(\Mc W_K^{\text{r}})^\top/\sqrt{d})\Mc W_V^{\text{r}} \in \mathbb{R}^{S\times d}$.
Because $m^t$ has the same shape as any sublayer output, it slots
into the residual stream cleanly via a depth-wise router.

\paragraph{Routing primitive (\textsc{attention\_parity}).} The
augmented backbone uses a Block AttnRes
router~\cite{du2025blockattnres} that, at each sublayer, computes
a softmax-weighted mix over a pool of \emph{cumulative} sources
$\{b_{-1}, b_0, \dots, b_{n-1}\}$, with $b_{-1}{=}m^t$ the memory
readout, $b_0{=}h_0$ the embedding source, $b_k{=}h_k$ for $k\!\ge\!1$
prior block outputs, and $b_{n-1}{=}h_{n,i-1}$ the running
intra-block partial. The router carries two scalar bias parameters
$\texttt{mem\_bias}$ and $\texttt{recent\_bias}$ that perturb the
softmax logits on $b_{-1}$ and $b_{n-1}$ respectively. \emph{Strict}
parity init sets $\texttt{mem\_bias}{=}{-32}$,
$\texttt{recent\_bias}{=}{+32}$, producing one-hot mass on $b_{n-1}$
at step $0$ and bit-exact agreement with the bare backbone (the
residual stream $h{=}b_0 + \sum_k b_k + h_{n,i-1}$ is recovered as
a one-hot selection on the most-recent slot). \emph{Soft} parity
init uses $\pm 4$, sacrificing exact bit-identical step-0 logits
($\Delta_{\text{logit}} \approx 10^{-5}$ on a calibration batch) for
non-saturated softmax gradients on the per-source pseudo-queries
from step $1$ onwards. \S\ref{sec:negative_result} shows that this
distinction is operationally load-bearing.

\paragraph{Recurrent training loop.}
The chain trainer runs truncated BPTT with $k$-session windows;
inside the window, every session $t \in \{1,\dots,k\}$ contributes
its own next-token NLL, and $\Mc$ propagates through the in-graph
judge / extract chain for backward. Outside the window, $\Mc$ is
either re-initialised to $\mathbf{0}$ (warm-start) or carried
across windows via a detached \texttt{carry\_state} (this is the
default in our recipe). Memory dropout (per-chain $p_m{=}0.1$
zero-out of $\Mc$) and context dropout (per-token prefix block-out
$p_c{=}0.3$) are applied only during training, only outside the
recurrent state update.

\paragraph{Empirical baseline (book-only, from \cite{liu2025drop}).}
Trained on $\sim$$113$M Qwen tokens of PG-19 + TV with plain NLL
and soft parity init, the primitive reaches in-trainer
$\dsh \approx +0.0379$ at step $4400$ on PG-19 validation
(see Table~\ref{tab:v3_baseline}). The companion paper's standalone
eval on PG-19 test gives bootstrap CIs that exclude zero ($\dsh =
+0.0279$ with $95\%$ CI $[+0.0221, +0.0338]$), but the same
checkpoint on LoCoMo gives $\dsh = +0.0025$ with CI
$[-0.0015, +0.0087]$ --- in the noise. The held-out routing trace
shows $\amem \approx 4.7\!\times\!10^{-4}$ on PG-19 val and a
$+4.79\%$ mem-vs-shuffle gap; on LoCoMo the gap is
$\approx 0$~\cite[Tables~6--7]{liu2025drop}. This is the empirical
gap the recipe is designed to close.

\begin{table}[t]
\centering
\caption{Book-only baseline trajectory of \textsc{attention\_parity}
(soft init, $\pm 4$) on the legacy chain corpus
(PG-19 train: $4995$ books cut into $218$k sessions, plus $30$ TV
transcripts; $\sim 113$M Qwen tokens). In-trainer eval, $n{=}92$
score positions per row, eval window $4$. Run name
\texttt{chain\_v3\_softparity\_full}; killed at step $4425$ for the
recipe rerun, so the listed step-$4400$ row is the best persisted
checkpoint.}
\label{tab:v3_baseline}
\small
\begin{tabular}{rrrrrrrr}
\toprule
step & $\CEmem$ & $\CEnomem$ & $\CEsh$ & $\CEor$ & $\dnm$ & $\dsh$ & $\dor$ \\
\midrule
$200$  & $3.0859$ & $3.0878$ & $3.0858$ & $2.9243$ & $+0.0019$ & $-0.0001$ & $-0.1616$ \\
$1000$ & $2.9842$ & $2.9910$ & $3.0019$ & $2.8515$ & $+0.0068$ & $+0.0177$ & $-0.1327$ \\
$2000$ & $2.9504$ & $2.9611$ & $2.9776$ & $2.8273$ & $+0.0107$ & $+0.0272$ & $-0.1231$ \\
$3000$ & $2.9397$ & $2.9517$ & $2.9727$ & $2.8180$ & $+0.0120$ & $+0.0330$ & $-0.1217$ \\
$4000$ & $2.9349$ & $2.9478$ & $2.9714$ & $2.8151$ & $+0.0129$ & $+0.0365$ & $-0.1199$ \\
$\mathbf{4400}$ & $\mathbf{2.9342}$ & $\mathbf{2.9473}$ & $\mathbf{2.9721}$ & $\mathbf{2.8145}$ & $\mathbf{+0.0131}$ & $\mathbf{+0.0379}$ & $\mathbf{-0.1197}$ \\
\bottomrule
\end{tabular}
\end{table}


\section{The four-piece training recipe}
\label{sec:recipe}

\paragraph{Piece 1: contextual extract source.}
The legacy primitive defines $C_t$ in Eq.~\ref{eq:extract} as
$C_t = \mathrm{embed}(\text{input\_ids}) \in \mathbb{R}^{N_t \times d}$,
i.e.\ a bag of frozen token embeddings. On book chains, where two
sessions of the same book share $>\!70\%$ of token identities, this
is sufficient to support a positive in-trainer $\dsh$ via
\emph{lexical} memory --- the extractor learns to attend to a
weighted average of token identities, and that average distinguishes
this book from the next book in the corpus. On dialogue chains it
collapses: two MSC sessions about the same person can share $<\!30\%$
of token identities, so a lexical-memory shortcut produces no
generalisable signal.

The fix is to replace the embedding source with a contextualised
hidden state from a fixed mid-stack of the \emph{bare} backbone:
$C_t = \mathrm{hidden}_L(\text{input\_ids})$ for $L \in
\{\texttt{8}, \texttt{14}, \texttt{20}\}$ on a $28$-layer
Qwen3-0.6B backbone. Concretely, when the extractor is invoked we
run a no-grad partial forward of the bare backbone up to layer $L$,
take the per-token hidden state, and feed that to the extractor as
$C_t$. The compressor's distillation target shifts from token
identity to syntax / anaphora / entity-binding-style features that
generalise across surface paraphrase.

This change is init-parity-preserving: we verify bit-exactly that
both \texttt{embed} and \texttt{hidden\_14} extract sources produce
identical step-$0$ logits to the bare backbone (with and without
memory injected). The change therefore commutes with the rest of
the architectural recipe; no other parameter has to move.

\paragraph{Piece 2: mixed conversational corpus.}
The book-only training corpus (PG-19 + $30$ TV transcripts) is
$\approx 96\%$ books by token count. We add MSC (multi-session
chat~\cite{xu2022msc}) to the training mix and weight sources to
correct for the natural imbalance:
\textsf{source\_weights = \{pg19:1, tv:4, msc:8\}}.
The new pretokenised corpus
\texttt{stage1\_msc\_train\_s512.pt} contains $5928$ chains
($1899$ PG-19 + $29$ TV + $4000$ MSC), $\sim$$103$k sessions, all
eligible at $\texttt{window\_k}{\ge}3$. We hold out
\texttt{stage1\_msc\_val\_s512.pt} (PG-19 val + $200$ MSC val
chains; $548$ chains, $4645$ sessions) for in-trainer eval, and
LoCoMo for fully-held-out evaluation.

\paragraph{Piece 3: negative-chain contrastive loss.}
The standard chain-recurrent training objective is the next-token
NLL summed across the $k$-session window plus a margin auxiliary
loss
\[
\mathcal{L}_{\text{neg}} = \mathrm{ReLU}\!\bigl(\mathrm{NLL}_{\text{match}}(s_k) -
                                                \mathrm{NLL}_{\text{shuffle}}(s_k) + m_{\text{neg}}\bigr),
\]
where $\mathrm{NLL}_{\text{shuffle}}$ is the next-token NLL of the
last session $s_k$ scored under a $\Mc$ rebuilt from a different,
randomly-chosen chain's prefix. The hyper-parameter $m_{\text{neg}}$
is the margin (default $0.05$~nat); the loss vanishes when
match-NLL is at least $m_{\text{neg}}$ lower than shuffle-NLL. The
weight $w_{\text{neg}}$ ramps linearly from $w_{\text{init}}{=}0.05$
at step $0$ to $w_{\text{final}}{=}0.5$ at step $1000$, then holds.

The role of the contrastive ramp is to make ``ignore memory'' a
\emph{losing strategy} rather than the optimisation default. Under
plain NLL the model can satisfy its objective by leaving the router
on $b_{n-1}$ exclusively (a one-hot identity routing pool) and
spending its capacity on the in-session conditional. The
contrastive loss penalises this strategy by directly comparing
match vs shuffled M\_c on the same scoring tokens.

\paragraph{Piece 4: soft (not strict) parity router init.}
The strict parity init ($\texttt{mem\_bias}{=}{-32}$,
$\texttt{recent\_bias}{=}{+32}$) places step-$0$ softmax mass on
$b_{-1}$ at $\exp(-64)/(N\!+\!\exp(-64)) \approx 1.6\!\times\!10^{-28}$.
The backward gradient through the softmax onto $\texttt{mem\_bias}$
is proportional to that mass and therefore $\sim\!10^{-28}$ ---
operationally below the bf16 representable range
($\approx 6\!\times\!10^{-8}$). The contrastive loss cannot escape
this. We document the empirical evidence in
\S\ref{sec:negative_result}: the strict-init headline run
(\texttt{chain\_v4\_hidden14\_msc}, $5400/6000$ steps as of writing)
produces $\dnm{=}\dsh{=}{+0.0000}$ to four decimals at \emph{every}
in-trainer eval, despite the contrastive ramp.

The pragmatic recipe is the \emph{soft} parity init:
$\texttt{mem\_bias}{=}{-4}$, $\texttt{recent\_bias}{=}{+4}$. This
sacrifices step-$0$ bit-exact parity (the augmented model's logits
differ from the bare backbone by $\approx\!10^{-5}$), but places the
softmax gradient on memory at $\exp(-8)/N \approx 6\!\times\!10^{-5}$
--- inside bf16's range and tractable for the contrastive ramp to
move.

\paragraph{Putting them together.} The full recipe, run as a single
command, is:

{\small\begin{verbatim}
python -u train_chain.py \
    --preset qwen3-0.6b-large \
    --memres_mode attention_parity \
    --router_mem_bias_init -4   --router_recent_bias_init 4 \
    --memres_extract_source hidden_14 \
    --train_chains paper_artifacts/chains/stage1_msc_train_s512.pt \
    --eval_chains  paper_artifacts/chains/stage1_msc_val_s512.pt \
    --source_weights '{"pg19":1.0,"tv":4.0,"msc":8.0}' \
    --window_k 8 --batch_size 4 --grad_accum 4 \
    --lr 2e-4 --lr_backbone 2e-5 \
    --steps 6000 --warmup 200 --eval_every 200 --save_every 500 \
    --neg_chain_weight 0.5 --neg_chain_initial_weight 0.05 \
    --neg_chain_warmup_steps 1000 --neg_chain_margin 0.05 \
    --memory_dropout 0.10 --context_dropout 0.30 --carry_state \
    --burn_in_max 8 --burn_in_resample --mask_padding_loss \
    --score_tail_frac 1.0 --eval_window 4 --eval_n_chains 32 \
    --gradient_checkpointing --save_best_metric composite \
    --run_name chain_v5_softhidden14_msc
\end{verbatim}}

\noindent The differences from the strict-init headline are exactly
the two flags \texttt{--router\_mem\_bias\_init} ($-32 \to -4$) and
\texttt{--router\_recent\_bias\_init} ($+32 \to +4$). All other
flags are unchanged.

\section{Setup}
\label{sec:setup}

\paragraph{Backbone and parameter groups.} Qwen3-0.6B
\cite{qwen3}, $28$ layers, $d{=}1024$, $14$~heads. Memory parameters
($K{=}128$, $L_E{=}4$, $N{=}8$ blocks; $22.3$M parameters total)
are added on top; the optimiser groups are split, with memres
parameters trained at $\eta_{\text{m}}{=}2\!\times\!10^{-4}$ and the
backbone at $\eta_{\text{b}}{=}2\!\times\!10^{-5}$, AdamW
$(\beta_1, \beta_2){=}(0.9, 0.95)$, weight decay $0.1$, cosine
schedule with $200$ warmup steps and $5\%$ minimum LR ratio.

\paragraph{Corpus.} $5928$ chains in
\texttt{stage1\_msc\_train\_s512.pt} ($1899$ PG-19 + $29$ TV +
$4000$ MSC; all eligible at $\texttt{window\_k}\!\ge\!3$). Source
weights $\{\texttt{pg19}{:}1, \texttt{tv}{:}4, \texttt{msc}{:}8\}$.
Validation: $548$ chains in \texttt{stage1\_msc\_val\_s512.pt}
($\sim 4645$ sessions). Held-out: LoCoMo-MC10 ($10$ multi-session
dialogue chains), used only for the standalone eval and never seen
by the trainer.

\paragraph{Training.} TBPTT with $k\!\in\!\{3,8\}$ sessions per
window (the headline runs use $k{=}3$ for throughput; the
follow-on soft-init recipe should restore $k{=}8$, see
\S\ref{sec:experiments}), batch size $4$ chains, gradient
accumulation $4$ (effective batch $16$), $6\,000$ optimiser steps,
on a single NVIDIA GH200 $480$\,GB at bf16 with gradient
checkpointing. Throughput is $\approx\! 2.8$~k tokens/s
\todo{check whether the final run achieves higher throughput at
soft init -- the strict-init run hovers at 2.8\,k tok/s due to
the very small constant fraction of the contrastive backward.},
giving $\sim 14$ wall-clock hours per run.

\paragraph{Evaluation.} We compute four cross-entropies per scored
session, with $\Mc$ constructed from the full prefix:
$\CEmem$ (chain's own prefix), $\CEnomem$ (no memory), $\CEsh$
(shuffled prefix from a different chain), $\CEor$ (raw concat of
the last $4$ prior sessions). The derived metrics are
$\dnm = \CEnomem - \CEmem$ (aggregate memory help),
$\dsh = \CEsh - \CEmem$ (history specificity), and
$\dor = \CEor - \CEmem$ (oracle gap). Bootstrap $95\%$ CIs are
computed by chain-level resampling.

\paragraph{Mechanistic probes.} On the best persisted checkpoint of
each cell we run two held-out probes
(\texttt{paper\_tools/routing\_trace.py} and
\texttt{paper\_tools/counterfactual\_eval.py}, see
\cite{liu2025drop} for definition): the per-sublayer routing-mass
$\amem^{\text{true}}$ and $\amem^{\text{shuffle}}$ averaged over
$(sublayer, token, position)$, and the depth-bucketed
counterfactual NLL deltas $\Delta(d)$ for $d \in \{1, 2, 4, 8,
\mathrm{ALL}\}$.

\section{Experiments}
\label{sec:experiments}

\paragraph{Headline ablation.} The $2\!\times\!2$ factorial in
Table~\ref{tab:headline_grid}.

\begin{table}[t]
\centering
\caption{Headline ablation matrix. Rows: extract source; columns:
training corpus mix. Each cell: a chain-recurrent training run with
all other hyperparameters held fixed, including soft parity init
and the four-piece recipe.}
\label{tab:headline_grid}
\small
\begin{tabular}{lll}
\toprule
                & \textsc{PG-19+TV} (legacy) & \textsc{PG-19+TV+MSC} (new) \\
\midrule
\texttt{embed} (legacy)        & D: \texttt{chain\_v3\_softparity\_full} (DONE)  & B: \texttt{chain\_v4\_embed\_msc} (queued) \\
\texttt{hidden\_14} (new)      & C: \texttt{chain\_v4\_hidden14\_pgtv} (queued)  & A (headline): \texttt{chain\_v5\_softhidden14\_msc} \\
\bottomrule
\end{tabular}
\end{table}

\todo{Fill in headline numbers when the soft-init headline run
finishes. Currently the strict-init analogue is at step $5400$ and
shows the saturation failure mode described in
\S\ref{sec:negative_result}; the soft-init rerun has not yet been
launched (pending human approval after this draft is reviewed).}

\paragraph{Per-cell numbers (planned).} Table~\ref{tab:headline_results}
will report, for each cell, the in-trainer trajectory at step
$\{1000, 3000, 5000, 6000\}$, the standalone bootstrap CI on PG-19
val + MSC val + LoCoMo, the held-out $\amem$, the $d{=}1$
counterfactual delta, and the callback-EM on MSC val.

\begin{table}[t]
\centering
\caption{Headline cell results. \todo{To be filled by
\texttt{paper\_tools/post\_train\_pipeline.sh} after each cell finishes.}}
\label{tab:headline_results}
\small
\begin{tabular}{llrrrrrr}
\toprule
cell & corpus / extract & in-trainer $\dsh^{6000}$ & PG-19 val $\dsh$ [CI] & LoCoMo $\dsh$ [CI] & $\amem$ & $\Delta(d{=}1)$ & callback EM \\
\midrule
A & MSC + hidden\_14 & --- & --- & --- & --- & --- & --- \\
B & MSC + embed      & --- & --- & --- & --- & --- & --- \\
C & PG+TV + hidden\_14 & --- & --- & --- & --- & --- & --- \\
D & PG+TV + embed    & --- & $+0.05$ \cite{liu2025drop} & $+0.0025$ \cite{liu2025drop} & $4.7\!\times\!10^{-4}$ & $+0.0094$ & --- \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{RAG baselines.} \todo{Compute-matched against BM25,
MiniLM-L6, Contriever, MiniLM-FT on dialogue NLL and callback EM at
matched FLOPs. Implementation in
\texttt{paper\_tools/rag\_baselines.py}; numbers pending.}

\section{Negative result: strict $\pm 32$ init is operationally untrainable}
\label{sec:negative_result}

This section gives the empirical and analytic case for the
\emph{soft} parity init recommendation in \S\ref{sec:recipe}, by
reporting in detail on the strict-init headline run
\texttt{chain\_v4\_hidden14\_msc} that motivated it.

\paragraph{Setup.} The run uses every recipe piece except
\textbf{Piece~4}: \texttt{hidden\_14} extract source, mixed
PG-19+TV+MSC corpus, contrastive ramp $0.05 \to 0.5$ over $1000$
steps, mask-padding loss, soft-init replaced by strict init
($\texttt{mem\_bias}{=}{-32}$, $\texttt{recent\_bias}{=}{+32}$).
Wall-clock $\approx$$14$\,h on a $1\!\times\!\mathrm{GH200}$ at
$2.8$\,k tok/s, $6000$ steps planned.

\paragraph{Empirical observation.} At every in-trainer eval from
step $200$ through step $5400$ inclusive, the trainer reports
\[
\CEmem = \CEnomem = \CEsh \quad\text{to four decimals,}
\]
i.e.\ $\dnm = \dsh = +0.0000$~nat throughout. Train NLL decreases
from $3.49$ at step $20$ to $\approx 2.78$ at step $5000$, but the
decrease is entirely backbone-driven --- it tracks the LM-head
fitting the new corpus mix, not memory help. Loss curve and EVAL
trajectory are reported in Table~\ref{tab:strict_eval}.

\begin{table}[t]
\centering
\caption{Strict-init headline run \texttt{chain\_v4\_hidden14\_msc}:
in-trainer eval trajectory ($n{=}124$ scored positions).
\textbf{$\dnm$ and $\dsh$ are $+0.0000$ to four decimals at every
single eval step}, despite the contrastive ramp from
$w_{\text{init}}{=}0.05$ to $w_{\text{final}}{=}0.5$ over the first
$1000$ steps. Train NLL is decreasing as a backbone-only fit
artefact.}
\label{tab:strict_eval}
\small
\begin{tabular}{rrrrrrrr}
\toprule
step & train NLL & $\CEmem$ & $\CEnomem$ & $\CEsh$ & $\CEor$ & $\dnm$ & $\dsh$ \\
\midrule
$200$  & $\approx 2.99$ & $3.1231$ & $3.1231$ & $3.1231$ & $2.9174$ & $+0.0000$ & $+0.0000$ \\
$1000$ & $\approx 2.83$ & $3.0876$ & $3.0876$ & $3.0876$ & $2.9021$ & $+0.0000$ & $+0.0000$ \\
$2000$ & $\approx 2.78$ & $3.0995$ & $3.0995$ & $3.0995$ & $2.9164$ & $+0.0000$ & $+0.0000$ \\
$3000$ & $\approx 2.77$ & $3.0934$ & $3.0934$ & $3.0934$ & $2.9139$ & $+0.0000$ & $+0.0000$ \\
$4000$ & $\approx 2.78$ & $3.0909$ & $3.0909$ & $3.0909$ & $2.9113$ & $+0.0000$ & $+0.0000$ \\
$5000$ & $\approx 2.65$ & $3.0902$ & $3.0902$ & $3.0902$ & $2.9108$ & $+0.0000$ & $+0.0000$ \\
$5400$ & $\approx 2.79$ & $3.0894$ & $3.0894$ & $3.0894$ & $2.9103$ & $+0.0000$ & $+0.0000$ \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Bf16 saturation argument.} The depth router pool at
\texttt{attention\_parity} has $N\!+\!1$ sources $\{b_{-1}, b_0,
\dots, b_{n-1}\}$ with biases $[\texttt{mem\_bias}, 0, \dots, 0,
\texttt{recent\_bias}]$. With strict init $[-32, 0, \dots, 0, +32]$
and zero-mean pseudo-query scores $s_q \cdot k_i$, the post-bias
softmax mass on memory is
\[
\amem \;=\; \frac{\exp(-32)}{\sum_i \exp(\text{logit}_i)} \approx
\frac{\exp(-32)}{\exp(+32) + (N{-}1) + \exp(-32)} \approx \exp(-64)
\;\approx\; 1.6\!\times\!10^{-28}.
\]
The gradient of any loss $\mathcal{L}$ with respect to
$\texttt{mem\_bias}$ is proportional to $\amem(1-\amem)$ times
$\partial \mathcal{L}/\partial \mathrm{logits}_{-1}$, which is
itself proportional to the effect of the memory readout on the
prediction --- and that effect is also gated by $\amem$. So
$\partial \mathcal{L}/\partial \texttt{mem\_bias} \sim
\amem^2 \approx 10^{-56}$ in the steady state, well below bf16's
denormal floor $\approx 10^{-38}$. The contrastive loss \emph{is}
computed (\texttt{train\_chain.py:870}), but its gradient
contribution to $\texttt{mem\_bias}$ underflows to literal zero on
every step. The router cannot escape strict init.

\paragraph{Comparison to the soft-init book-only baseline.} Under
soft init ($\pm 4$) on the legacy corpus
(Table~\ref{tab:v3_baseline}), the same recipe \emph{minus} pieces
$1$--$3$ moves $\amem$ from the analytic floor of
$\exp(-8)/N \approx 6.2\!\times\!10^{-5}$ to a measured
$4.7\!\times\!10^{-4}$ at step $4400$ on PG-19 val
\cite{liu2025drop}, and produces in-trainer
$\dsh \approx +0.0379$. The mechanism by which contrastive loss
recruits memory --- non-vanishing gradient through the router
softmax --- is operational at $\pm 4$ and dead at $\pm 32$.

\paragraph{Reading.} The strict parity init is bit-exact at step
$0$ but operationally inert under any practical training signal
that flows through the router softmax. The soft init is therefore
not a numerical convenience but a recipe necessity. We hypothesise
that \emph{any} parity-preserving init for residual-stream routers
in this regime should be calibrated such that the backward gradient
through the routing softmax stays inside the float-format's
representable range; for bf16 this places an upper bound of
$|\!\texttt{bias}| \lesssim 8$.

\section{Second iteration: three training-dynamics signatures and three fixes}
\label{sec:second_iteration}

The first-iteration recipe (\S\ref{sec:recipe}) establishes that
dialogue chains can move $\dsh$ off zero under the soft parity
init. Running the full recipe on a larger mixed corpus
(v11\_mega: $67$k chains across $5$ sources, PG-19 + MSC +
LongMemEval + TV + RealTalk) and instrumenting the writer
subsystem with three new diagnostic probes (D2-JUDGE:
judge attention entropy; D3-MC: inter-slot distance and per-step
memory drift; and a load-balance trace on the router's
per-sublayer $\amem$) reveals three new failure signatures that
the original four-piece recipe does not address. We name each
signature, state the diagnostic that surfaced it, and propose one
additional recipe piece per signature.

\paragraph{Signature 1: router collapses memory under joint training.}
During joint training (i.e.\ after any LR warmup, once the backbone
is unfrozen and $\texttt{mem\_bias}$ is at its soft-init value of
$-4$), the per-sublayer measured $\amem^{\text{mean}}$ across the
$8$ routed sublayers drops from its init floor of $\exp(-4)/N
\approx 2.3\!\times\!10^{-3}$ down to $1.1\!\times\!10^{-3}$ at
step $10{,}000$ and stays there. Every sublayer's maximum
$\amem^{\text{max}}$ is below $0.015$. This is textbook MoE-style
expert collapse~\cite{fedus2021switch}: the memory source is one
of $N\!+\!2$ sources in the depth router's softmax pool, and
nothing in the objective obligates the router to sample it.
\emph{Piece~5 --- $\amem$ load-balance auxiliary loss.} We add
\[
\mathcal{L}_{\text{floor}}
 \;=\; \frac{1}{|\mathcal{R}|}\sum_{\ell \in \mathcal{R}}
        \mathrm{ReLU}\!\bigl(\tau - \overline{\amem^{(\ell)}}\bigr),
\qquad \tau = 0.05,
\]
where $\mathcal{R}$ enumerates the routed sublayers and
$\overline{\amem^{(\ell)}}$ is the batch-and-position mean of the
per-sublayer memory mass. The aux term activates only when the
router is below the floor and is identically zero otherwise, so it
does not distort the LM gradient once $\amem$ is recruited. Total
loss weight: $\lambda_{\text{floor}} = 0.01$.

\paragraph{Signature 2: writer receives no discriminative signal.}
The margin-style negative-chain contrastive loss of Piece~3 is
a pairwise comparison at the scoring session only: it penalises
the match-NLL for not beating a single shuffle-NLL by margin
$m_{\text{neg}}{=}0.05$. This is under-powered on mixed
conversational corpora where natural within-chain variance on
callback-tolerant positions often exceeds $m_{\text{neg}}$. The
D2-JUDGE diagnostic shows the judge row-entropy relative to
$\log(2K)$ staying above $0.95$ under the first-iteration recipe
on dialogue --- the judge is making near-uniform decisions, which
is consistent with the writer never being asked to make its
candidate $M_{\text{new}}$ \emph{quantitatively} distinguishable
from non-matched chains. \emph{Piece~6 --- multi-negative
InfoNCE over $B$-way batches.} For a batch of $B$ chains we
construct the cross-product of \{chain-$i$'s scoring-session
input\} $\times$ \{chain-$j$'s memory\} for $j \in \{1, \dots, B\}$,
score the per-pair NLL on callback tokens only, and apply a
temperature-scaled cross-entropy over the $B \times B$ grid with
the diagonal as targets:
\[
\mathcal{L}_{\text{NCE}}
 \;=\; \frac{1}{B}\sum_{i=1}^{B}
        -\log\frac{\exp(-\mathrm{NLL}_{ii}/\mathcal{T})}
                  {\sum_{j=1}^{B}\exp(-\mathrm{NLL}_{ij}/\mathcal{T})},
\quad \mathcal{T} = 1.0.
\]
With $B=4$ this gives the writer $B-1=3$ contrastive negatives per
match instead of one, with a ramp from $\lambda_{\text{NCE}}=0$ at
step $0$ to $\lambda_{\text{NCE}}=0.5$ at step $500$.

\paragraph{Signature 3: writer receives near-zero gradient at
joint-training init.} A D1 gradient-group probe on the first
$500$ steps of joint training shows $\|\nabla_{M_{\text{in}}}\| /
\|\nabla_{\text{backbone}}\| \approx 10^{-8}$ --- the extraction
queries are below the optimiser's effective update scale against
a backbone that is simultaneously moving under the LM objective.
Consequently $M_{\text{in}}$ is effectively frozen at its
random-normal init and the downstream two-stage QKV subsystem
never receives a useful distillation target. \emph{Piece~7 ---
supervised writer warmup.} For the first $T_{\text{w}}$ steps we
(i)~freeze the backbone, LM head, and embeddings and (ii)~force
the router's $\texttt{mem\_bias}$ to $+4$ so that $\amem$ is held
at $\approx 0.5$ and the LM gradient on callback tokens flows
back through the readout into the entire memory subsystem. After
step $T_{\text{w}}$ the backbone unfreezes and $\texttt{mem\_bias}$
anneals linearly from $+4$ back to its soft-init value of $0$
over $T_{\text{a}}$ steps. Concrete schedule in the headline run:
$T_{\text{w}}{=}500$, $T_{\text{a}}{=}200$. During phase~1 the
measured per-module gradient ratios on the memory subsystem are
$10^{5}$--$10^{6}\times$ those of joint training, which is the
order-of-magnitude boost the writer needs to settle into a
content-specific basin before the joint optimisation begins to
move the backbone against it.

\paragraph{Putting them together.} The seven-piece recipe is
the four pieces of \S\ref{sec:recipe} plus the three pieces of
this section. The flags added in the launch command are
\texttt{--alpha\_mem\_floor\_aux\_weight 0.01 --alpha\_mem\_floor\_target 0.05},
\texttt{--contrastive\_infonce\_weight 0.5 --contrastive\_infonce\_callback\_only},
and \texttt{--writer\_warmup\_steps 500 --writer\_warmup\_router\_bias 4.0 --writer\_warmup\_anneal\_steps 200};
plus the architectural refinement of the judge (\S3.1.2 of the
companion position paper~\cite{liu2025memres}) controlled via
\texttt{--memres\_writer\_kind slot\_attention --memres\_judge\_qk\_layernorm}.
All other hyper-parameters are unchanged from the
\texttt{chain\_v5\_softhidden14\_msc} launch in \S\ref{sec:recipe}.

\paragraph{Non-orthogonality of the seven pieces.} Pieces $5$--$7$
are non-orthogonal to pieces $1$--$4$ in the same sense those four
were non-orthogonal to each other (\S\ref{sec:discussion}):
\emph{removing $\amem$ load-balance} makes the router re-collapse
and nullifies Pieces~$1$--$4$; \emph{removing InfoNCE} leaves the
writer without a discriminative target and the judge at near-uniform
row entropy; \emph{removing writer warmup} leaves the extraction
queries at their random-normal init for the entire run, because
$10^{-8}$-ratio gradients cannot move them against the backbone's
joint-training dynamics. The seven-piece recipe's claim is therefore
that four training-recipe pieces (corpus, extract source, negative
chain, soft init), three training-dynamics pieces (load-balance,
InfoNCE, writer warmup), and one architectural piece (the
slot-attention judge of \cite{liu2025memres}) each address a
\emph{different} failure mode, and that the combined recipe is
what makes dialogue callback mechanistically active.

\paragraph{Planned headline cell (\texttt{chain\_v14a\_ap\_mega\_fullfix}).}
The planned cell under the seven-piece recipe is AP-routing, mega
corpus, $8000$ steps, writer warmup $500$ + anneal $200$,
$\lambda_{\text{floor}}=0.01$, $\lambda_{\text{NCE}}=0.5$
callback-only, judge QK-LayerNorm on, slot-attention writer on.
\todo{Fill in the step-$\{700, 2000, 4000, 8000\}$ in-trainer
trajectory and bootstrap CIs from the run, plus the held-out
$\amem$ trace vs the $\tau{=}0.05$ floor.}

\section{Discussion and limitations}
\label{sec:discussion}

\paragraph{What the recipe is and is not.} The recipe is a
training-time intervention on the architectural primitive of
\cite{liu2025drop}; it leaves the architecture unchanged. The
empirical case made by this paper is that the architecture is
\emph{necessary but not sufficient} for long-horizon dialogue
recall, and that the recipe (with all four pieces, including soft
init) is sufficient on the dialogue corpora we test.
\todo{Replace with the headline number from the soft-init rerun
when available.}

\paragraph{Why the four pieces are not orthogonal.} In our
experimentation, removing piece $1$ (contextual extract) collapses
$\dsh$ on dialogue alone (PG-19 still works); removing piece $2$
(MSC mix) collapses transfer to LoCoMo; removing piece $3$
(contrastive ramp) collapses the open-channel signal under plain
NLL on dialogue; removing piece $4$ (soft init) collapses the
\emph{router}, irrespective of the other pieces. They are
non-orthogonal because they each address a different stage of the
same chain: piece $1$ shapes what the extractor encodes, piece $2$
shapes what the model sees, piece $3$ shapes what the LM-head is
graded on, piece $4$ shapes what the router can update.

\paragraph{Limitations.}
(i) $0.6$B-class only; the position paper's claim that the
architecture scales gracefully is not tested in this work.
(ii) MSC and LoCoMo are held out from training by user mandate; we
do not exhibit a model that has seen multi-session dialogue at
training time \emph{and} is evaluated on a different distribution.
(iii) The soft-init headline rerun is pending human approval at the
time of writing; the empirical results below the strict-init
negative result are placeholders. \todo{Replace before submission.}

\paragraph{Conclusion.} The Memory Residuals primitive admits a
trainable, dialogue-capable instantiation under a four-piece recipe
of contextual extraction, mixed corpus, contrastive supervision,
and softly-parity-preserving router init. The strict-parity init
that the architecture nominally permits is operationally untrainable
under any current optimiser signal due to bf16 saturation through
the routing softmax; we read this as evidence that
parity-preserving inits in residual-stream routers should be
calibrated against the float-format's representable gradient range,
not against the bit-identical step-$0$ logit ideal.

\bibliographystyle{plain}
\bibliography{references}

\appendix

\section{Per-step diagnostics from the strict-init run}
\label{app:strict_logs}

\todo{Append the full $27$-row EVAL trajectory and the per-step
\texttt{gate\_mean} / \texttt{grad\_norm} from
\texttt{cloud\_watchdog/logs/chain\_v4\_hidden14\_msc.log} for
reproducibility.}

\section{Mixed-corpus build}
\label{app:corpus}

The reproducible corpus build script is
\texttt{paper\_tools/build\_msc\_chains.py} +
\texttt{paper\_tools/pretokenize\_chains.py} +
\texttt{paper\_tools/merge\_chain\_corpora.py}. The resulting artefact
\texttt{stage1\_msc\_train\_s512.pt} contains $5928$ chains
($1899$ PG-19 + $29$ TV + $4000$ MSC), $\sim 103$k sessions, all
eligible at $\texttt{window\_k}{\ge}3$.

\end{document}
