% experiments/exp2_chain_recipe/section7_v14_findings.tex
%
% Replacement for Section 7 of draft.tex. Reports the v14 ablation
% series on the D4v2 synthetic persona-callback corpus, in which the
% three-piece writer warmup of the original Section 7 ("Second
% iteration") is shown to be empirically counterproductive.
%
% This file is meant to be \input'd in place of the original
% Section 7. It does not modify draft.tex.

\section{Third iteration: the writer-warmup catastrophe on a synthetic callback corpus}
\label{sec:third_iteration}

The seven-piece recipe of \S\ref{sec:second_iteration} was designed
against three diagnostic signatures that we surfaced on the
\texttt{v11\_mega} mixed corpus: router collapse under joint
training, judge near-uniformity, and a near-zero
extraction-query gradient at joint-training init. Two of those
fixes --- the $\amem$ load-balance auxiliary
($\lambda_{\text{floor}}{=}0.01$, $\tau{=}0.05$) and the
multi-negative InfoNCE term ($\lambda_{\text{NCE}}{=}0.5$,
callback-only) --- are robust under the third-iteration runs
reported here. The third fix, the supervised writer warmup
(forced $\texttt{mem\_bias}{=}{+4}$ over the first
$T_{\text{w}}{=}500$ steps with $T_{\text{a}}{=}200$ steps of
anneal), is not. We report a six-cell ablation matrix on a clean
synthetic callback corpus that pins this down to a specific failure
mode of the warmup itself, and to a residual imprint on the writer
that survives the anneal.

\paragraph{Why a synthetic callback corpus.}
\texttt{v11\_mega} is heterogeneous: PG-19, MSC, LongMemEval, TV,
RealTalk. Its strength as a recipe-development corpus is that any
recipe that survives it generalises; its weakness as a
\emph{diagnostic} corpus is that callback positions are
heuristically labelled and the per-chain variance on
$\dsh$-style metrics is high enough that small but real signals are
hard to separate from noise in a single-cell run. We therefore built
\textbf{D4v2}, a fully synthetic persona-callback corpus, as a
controlled probe for the writer subsystem in isolation. Each chain
has $9$ filler-or-evidence sessions plus a single callback session;
the evidence is \emph{double}, i.e.\ each chain places the
target persona fact in exactly $2$ of the $9$ prior sessions and
fills the remaining $7$ with on-distribution distractor sessions
generated by the same template machinery. The callback session
asks a question whose ground-truth answer is uniquely determined by
the evidence sessions and is invariant under any subset of
the filler sessions. We pretokenise $5000$ train and $500$ val
chains at $512$ tokens / session and score the callback target
under four memory regimes ($\Mc^{\text{match}}$, $\mathbf{0}$,
$\Mc^{\text{shuffle}}$, $\Mc^{\text{oracle-evidence-only}}$); the
primary metric is \emph{evidence\_lift}, the per-token NLL gap
between the match and zero-memory regimes on the evidence-bearing
target tokens of the callback session. By construction the corpus
admits a maximum evidence\_lift of $\approx +0.5$~nat and a clean
zero under any non-evidence-recruiting policy.

\paragraph{The v14 ablation matrix.} Six cells were run on D4v2
under the seven-piece recipe of \S\ref{sec:second_iteration}, with a
frozen Qwen3-0.6B backbone, the slot-attention judge with
QK-LayerNorm, the $\amem$ load-balance auxiliary
($\lambda_{\text{floor}}{=}0.01$, $\tau{=}0.05$), and the
callback-only InfoNCE
($\lambda_{\text{NCE}}{=}0.5$). The cells differ only in three
knobs: whether the extractor's contextual input is
RMS-normalised before the cross-attention
(\texttt{memres\_extract\_input\_norm}, see
\S\ref{sec:revised_recipe}); the magnitude of the warmup-time
forced $\texttt{mem\_bias}$; and whether warmup is applied at all.
Table~\ref{tab:v14_grid} lists the cells and outcomes.

\begin{table}[t]
\centering
\caption{The v14 ablation matrix on D4v2. All cells share frozen
Qwen3-0.6B, slot-attention judge with QK-LayerNorm, $\amem$
load-balance ($\tau{=}0.05$, $\lambda{=}0.01$), and callback-only
InfoNCE ($\lambda{=}0.5$). Cells differ only in the three knobs
shown; \emph{norm} abbreviates
\texttt{memres\_extract\_input\_norm}, \emph{wb} the warmup-phase
$\texttt{mem\_bias}$, and $T_{\text{w}}$ the warmup step count.
The v14i cell uses a reduced extract LR
$\eta_{\text{m}}{=}3{\times}10^{-5}$ (all others
$2{\times}10^{-4}$). \emph{evidence\_lift} is the per-token NLL
gap on evidence-bearing callback targets between the match-memory
and zero-memory regimes; positive means memory helps. ``killed''
means the run was halted because train NLL exceeded
$\log(|\mathcal{V}|) \approx 11.9$~nat for $\ge 500$ consecutive
steps during warmup.}
\label{tab:v14_grid}
\small
\begin{tabular}{lcccll}
\toprule
cell & norm & wb & $T_{\text{w}}$ & writer\_kind & best evidence\_lift \\
\midrule
v14g & on  & $+4.0$ & $500$ & slot\_attention      & killed @760 (LM destroyed) \\
v14h & off & $+4.0$ & $500$ & slot\_attention      & $\approx -0.05$ post-anneal (lock-in) \\
v14i & on  & $+4.0$ & $500$ & slot\_attention\,$^*$ & killed @1140 (recovers, then kills) \\
v14j & on  & $+2.0$ & $500$ & slot\_attention      & $-0.03$ to $-0.12$ post-anneal \\
\textbf{v14k} & \textbf{on}  & $\mathbf{0.0}$ & $\mathbf{0}$ & \textbf{slot\_attention} & $\boldsymbol{+0.13}$ \textbf{@2600 (ckpt @3000)} \\
v14l & on  & $+4.0$ & $500$ & slot\_attention\_full & not run (budget) \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{The warmup catastrophe.}
Cells v14g and v14i, both with the input-norm fix on and warmup
$\texttt{mem\_bias}{=}{+4}$, are killed during the warmup phase
itself: train NLL crosses $\log(|\mathcal{V}|)$ within
$\sim 60$ steps of warmup onset and stays above the
chance-prediction threshold for $500$--$700$ steps before slowly
recovering. The mechanism is mechanical and follows from the
seven-piece recipe's own design choices stacked. Forcing
$\texttt{mem\_bias}{=}{+4}$ with the existing
$\texttt{recent\_bias}{=}{+4}$ and zero-bias remaining sources
saturates the router softmax at $\amem \approx 0.98$ on the memory
slot at \emph{every} routed sublayer of the augmented backbone. With
the backbone frozen, the only learnable response to this regime is
the memory subsystem itself: $98\%$ of every sublayer's input is
the readout $m^t$, and the LM head is asked to produce next-token
logits from a residual stream that is almost entirely a learned
compression of $C_t$ rather than the bare backbone's representation.
The optimiser receives a $\amem^{(\ell)} \approx 0.98$ gradient
through the writer's
$W_K^{\text{ext}}, W_V^{\text{ext}}$ matrices that is empirically
$10^8$--$10^9$~unclipped per step (clipped to $\sim 10^{-9}$
post-norm, so the second-order Adam moments are immediately
saturated), and the LM head, frozen, cannot move to absorb the
mismatch. The net effect is that the LM is destroyed in the first
hundred steps of warmup; once the bias starts annealing toward $0$
at step $T_{\text{w}}$, the readout subsystem is trained to a
``compress everything'' fixed point that is mechanically inconsistent
with the bare backbone's expected residual-stream geometry.

\paragraph{Recent-bias lock-in.}
The intermediate cells v14h and v14j survive the warmup phase
intact --- v14h because the input-norm fix is \emph{off} and the
extract gradients never inflate, v14j because the warmup
$\texttt{mem\_bias}{=}{+2}$ saturates more weakly
($\amem \approx 0.88$) and so the writer never compresses
strongly enough to crash the LM. Both cells, however, exhibit a
follow-on failure that we did not anticipate from the v11
diagnostics. After the bias anneals back to $0$, the trained writer
has been imprinted with a ``compress everything'' policy that the
post-anneal optimisation cannot undo: under the joint objective
(NLL + InfoNCE + alpha-floor aux), the router converges to a
near-pure recent-bias regime where $\amem^{\text{mean}}$ stabilises
at the floor $\tau{=}0.05$ \emph{exactly} (the aux loss saturates
the inequality), and the readout $m^t$ contributes only the
weakest signal compatible with the floor. Evidence\_lift on D4v2
is consistently \emph{negative} in both cells across the entire
post-anneal trajectory ($-0.03$ to $-0.12$), i.e.\ the memory
subsystem is producing a \emph{worse} prediction on evidence
tokens than the no-memory baseline. The ratio of warmup magnitudes
($+4.0$ vs $+2.0$) does not separate the cells: the post-anneal
recent-bias lock-in is qualitatively the same. We read this as
evidence that the warmup phase, in the regime where the backbone
is frozen and the router has no other source of signal, trains the
writer toward an extract policy that is \emph{anti-aligned} with the
evidence-token gradient that the post-anneal joint training would
have selected.

\paragraph{The headline negative result.}
Cell v14k removes the warmup phase entirely
($T_{\text{w}}{=}0$, warmup $\texttt{mem\_bias}{=}0$); the router
starts from soft init ($\texttt{mem\_bias\_init}{=}0$) and is
immediately exposed to the full joint objective. Evidence\_lift
crosses zero from below at step $\sim 1100$, climbs to $+0.10$ by
step $2000$, and stabilises at $+0.13$ from step $2600$ onwards;
the best-checkpoint metric is persisted at step $3000$. Under the
seven-piece recipe \emph{minus} writer warmup, with the input-norm
fix on, the alpha-floor and InfoNCE pieces in place, and the
slot-attention judge unchanged, the writer recruits memory to a
sustained positive evidence\_lift on a clean synthetic callback
corpus. The headline empirical finding of this section is therefore
the \emph{negative} one against \S\ref{sec:second_iteration}'s
Piece~$7$: the supervised writer warmup, controlled by the three
\texttt{--writer\_warmup\_*} flags of \S\ref{sec:second_iteration},
was \emph{not} a necessary fix to the near-zero extraction-query
gradient at joint-training init. On D4v2 it is at best inert (where the warmup is destructive
enough to crash the run and force a kill) and at worst silently
counterproductive (where the warmup completes cleanly, the LM
recovers post-anneal, and the writer is left in a recent-bias
fixed point that drives evidence\_lift below zero).

\paragraph{Reading and scope.}
We are deliberately careful about the scope of this finding. The
v14 cells run on a single corpus (D4v2), with a single backbone
(frozen Qwen3-0.6B), and a single architectural realisation
(slot-attention writer + judge QK-LayerNorm, AP routing, soft init
$\pm 4$). What we are claiming is that on this corpus and at this
scale, removing the writer warmup of \S\ref{sec:second_iteration}
strictly improves the headline metric, and that the alternative
recipe --- input-norm + alpha-floor + InfoNCE, no warmup --- is the
first recipe in this paper that produces a sustained positive
evidence\_lift on a controlled callback corpus. We do not claim
that writer warmup is harmful in regimes that differ in any of
those three axes, and the v15 series is designed to probe each
axis in turn (Table~\ref{tab:v15_pending}).

\begin{table}[t]
\centering
\caption{Companion v15 cells (currently running) that probe the
three axes along which the v14k recipe is conservative: backbone
training, extract source, and scale. All v15 cells use the
\S\ref{sec:revised_recipe} three-piece recipe (input-RMSNorm,
no warmup, alpha-floor $+$ InfoNCE) and differ only in the listed
knob. v15a is expected to land at $+0.10$ to $+0.15$
evidence\_lift, matching v14k.}
\label{tab:v15_pending}
\small
\begin{tabular}{llcl}
\toprule
cell & delta vs.\ v14k & evidence\_lift & status \\
\midrule
v15a & exact replicate                              & \todo{} & running \\
v15b & $+$ unfrozen backbone (lr $2{\times}10^{-5}$) & \todo{} & running \\
v15c & $+$ extract source $=$ \texttt{embed}         & \todo{} & running \\
v15e & $+$ Qwen3-1.7B, frozen                       & \todo{} & queued  \\
v15f & $+$ Qwen3-1.7B, joint-train                  & \todo{} & queued  \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Standalone evaluation (pending).}
Once v14k's persisted step-$3000$ checkpoint is taken through the
full standalone evaluation pipeline, we will report
$\dnm$, $\dsh$, $\dor$, and the depth-bucketed counterfactual
\emph{capture ratio} on three corpora: D4v2 val
($500$ chains, $2$ evidence + $7$ filler + $1$ callback), LoCoMo
($10$ multi-session conversations, $272$ sessions), and MSC test
($500$ chains, $\sim 2500$ sessions). The placeholder structure for
those numbers is in Table~\ref{tab:v14k_standalone}; the missing
entries are gated on the v14k post-train pipeline run.

\begin{table}[t]
\centering
\caption{Planned standalone evaluation of the v14k step-$3000$
checkpoint on the three target corpora. \emph{capture ratio}
is the depth-bucketed
$\Delta(d{=}\mathrm{ALL}) / \Delta(d{=}\mathrm{ORACLE})$ on the
counterfactual probe.}
\label{tab:v14k_standalone}
\small
\begin{tabular}{lrrrr}
\toprule
corpus & $\dnm$ & $\dsh$ & $\dor$ & capture ratio \\
\midrule
D4v2 val (500 chains)            & \todo{} & \todo{} & \todo{} & \todo{} \\
LoCoMo (10 conv, 272 sessions)   & \todo{} & \todo{} & \todo{} & \todo{} \\
MSC test (500 chains, $\sim$2.5k sessions) & \todo{} & \todo{} & \todo{} & \todo{} \\
\bottomrule
\end{tabular}
\end{table}
